---
title: "Codebook for Getting and Cleaning Data Course Project"
author: "Mark van den Broek"
date: "Thursday, October 15, 2015"
output: md_document
---
## Introduction
The purpose of this project is to demonstrate the ability to collect, work with, and clean a data set. The goal is to prepare tidy data that can be used for later analysis. As a result of this analysis, the input data are transformed to  two new tidy datasets. This Codebook describes the input data and the process to transform the data. 

The data that is analyzed comes from the are of wearable computing. This is one of the most exciting areas in all of data science right now. Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. The data represent data collected from the accelerometers from the Samsung Galaxy S smartphone. 
```{r}
```

## Data description
###Data Source
The data files that we will be using for this analysis can be found here: 

https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip 

A full description is available at the site where the data was obtained: 

http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones 

###Some background on the input data
Some background on the input data and how they were obtained and preprocessed: 

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

### More detail on the input data
The zip archive contains a number of files: data files and explanatory files. I refer to the explanatory files for more detail. In particular the readme.txt file and features_info.txt contain detailed information. 

I will now give a summary that is specific to this subsequent data analysis. The following files are used: 

Two files containing feature names and class labels respectively:

- 'features.txt': List of all features.
- 'activity_labels.txt': Links the class labels with their activity name.

Six files containing data, three training data sets and three test data sets. 

- 'train/X_train.txt': Training set.
- 'train/y_train.txt': Training labels.
- 'train/subject_train.txt': Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 
- 'test/X_test.txt': Test set.
- 'test/y_test.txt': Test labels.
- 'test/subject_train.txt': Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 

For this analysis we will be using a subset of the features in x_train (and x_test): namely only the features that correspond to a mean or standard deviation of the original data measurements. 

A few last notes: 

- Features are normalized and bounded within [-1,1].
- Each feature vector is a row on the text file.

###Feature definitions
The following was taken from the file features_info.txt and gives detailed information on the individual features in the data:

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

- tBodyAcc-XYZ
- tGravityAcc-XYZ
- tBodyAccJerk-XYZ
- tBodyGyro-XYZ
- tBodyGyroJerk-XYZ
- tBodyAccMag
- tGravityAccMag
- tBodyAccJerkMag
- tBodyGyroMag
- tBodyGyroJerkMag
- fBodyAcc-XYZ
- fBodyAccJerk-XYZ
- fBodyGyro-XYZ
- fBodyAccMag
- fBodyAccJerkMag
- fBodyGyroMag
- fBodyGyroJerkMag 

The set of variables that were estimated from these signals are: 

- mean(): Mean value
- std(): Standard deviation
- mad(): Median absolute deviation 
- max(): Largest value in array
- min(): Smallest value in array
- sma(): Signal magnitude area
- energy(): Energy measure. Sum of the squares divided by the number of values. 
- iqr(): Interquartile range 
- entropy(): Signal entropy
- arCoeff(): Autorregresion coefficients with Burg order equal to 4
- correlation(): correlation coefficient between two signals
- maxInds(): index of the frequency component with largest magnitude
- meanFreq(): Weighted average of the frequency components to obtain a mean frequency
- skewness(): skewness of the frequency domain signal 
- kurtosis(): kurtosis of the frequency domain signal 
- bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each window.
- angle(): Angle between to vectors.

Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

- gravityMean
- tBodyAccMean
- tBodyAccJerkMean
- tBodyGyroMean
- tBodyGyroJerkMean

The complete list of variables of each feature vector is available in 'features.txt'

## Reading the input data
We have already downloaded the zip file and extracted it in our working directory. All data are in a subfolder. 

###Reading the activity names and feature names
Now, we will read the activity names first. 
```{r}
wd <- getwd()
datapath <- file.path(wd, "UCI HAR Dataset" )

filename <- "activity_labels.txt"
actNames <-read.table(file.path(datapath, filename),col.names = c("ActivityId", "ActivityName"))
```

Next, we are going to read the feature names: 
```{r}
filename <- "features.txt"
features <-read.table(file.path(datapath, filename),col.names = c("VarId", "VarName"))
featureNames <- as.character(features$VarName)
```

We will now select the subset of features that correspond to a mean or st. deviation. We do this by using regular expressions applied to the feature names: 
```{r}
meanFNames <- featureNames[sapply(featureNames, regexpr, pattern = "mean") > 0]
stdFNames <- featureNames[sapply(featureNames, regexpr, pattern = "std") > 0]
selectedFNames <- c(meanFNames , stdFNames)
```

We tidy up the names of the features by removing the "()" in the names:
```{r}
tidyFNames <- sub(pattern = "\\(\\)", replacement = "", x = selectedFNames)
length(tidyFNames)
```
We are now left with the following feature names: 
```{r}
tidyFNames
```

###Reading the test and training data
The training and test feature data are fixed width files with 16 cols per variable. In order to read only the relevant features, we prepare a width vector that contains positive and negative widths for variables that we want to read and skip respectively.  
```{r}
widthvector <- 16 * (2 *(featureNames %in% selectedFNames)-1)
```

Now we will read the three training files: 
```{r}
path <- file.path(datapath, "Train" )
filename <- "X_train.txt"
train_raw <- read.fwf(file.path(path, filename), width = widthvector, buffersize = 100, col.names = tidyFNames )
filename <- "y_train.txt"
train_y <- read.table(file.path(path, filename), col.names =c("ActivityId"))
filename <- "subject_train.txt"
train_subj <- read.table(file.path(path, filename), col.names = c("SubjectId"))
```

And we read the three test files: 
```{r}
path <- file.path(datapath, "Test" )
filename <- "X_test.txt"
test_raw <-  read.fwf(file.path(path, filename), width = widthvector, buffersize = 100, col.names = tidyFNames  )
filename <- "y_test.txt"
test_y <- read.table(file.path(path, filename), col.names =c("ActivityId"))
filename <- "subject_test.txt"
test_subj <- read.table(file.path(path, filename),col.names = c("SubjectId"))
```

We have given appropriate column names to all the variables: we have used the feature names that we created before and we have used ActivityId and SubjectId for the variables in the y and subject files respectively. 

## Merging the data
We will merge the data by first combining the features and the ActivityId and SubjectId variables for both test and train. Subsequently we merge the train and test data to one data set. This can be done in one statement: 

```{r}
totaldata <- rbind( cbind(train_raw, train_y, train_subj), cbind(test_raw, test_y, test_subj) )
```

We will now also join the activity name to the dataset: 
```{r}
totaldata <- merge(totaldata, actNames, by.x = "ActivityId" , by.y = "ActivityId")
```

## Creating the dataset with mean variables per activity and per subject
###Calculating the mean per subjectId and ActivityId
We are now in good shape to create a dataset that contains the mean of each feature, grouped by subjectId and by ActivityId. We will use the aggregate function in R and sort it afterwards. We will show the first 5 rows and columns.
```{r}
cols <- colnames(totaldata)
cols <- cols[!cols %in% c("ActivityName","SubjectId")] 
meanDataPerActivitySubject <- aggregate(x = totaldata[cols], 
                  by = list( ActivityName = totaldata$ActivityName, SubjectId = totaldata$SubjectId), 
                  FUN = mean)

meanDataPerActivitySubject <- meanDataPerActivitySubject[order(meanDataPerActivitySubject$ActivityName, 
                                                               meanDataPerActivitySubject$SubjectId),]
meanDataPerActivitySubject[1:5][1:5,]
```

###Writing the tidy datasets to files
We will now write the two tidy datasets to files: 
```{r}
filename <- "tidyMeanData.txt"
write.table(x = meanDataPerActivitySubject , filename, quote = F, row.names = F )

filename <- "tidyTotalData.txt"
write.table(x = totaldata , filename, quote = F, row.names = F )
```